Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2010-06-11 Text/migration.tex
Dick's changes 6/11/10
git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@65 3b7f7548-5c66-0410-83ab-3997167635fc

Section 2 below presents our numerical methods and initial conditions.  We describe and discuss the simulation results in Section 3, and summarize our main conclusions in Section 4.

@@

%Scott: A number of things in this paragraph need to verified by further analysis.
{\it The $t = 0$ Planet Run.} {\bf See note in Latex file.} Figure 1 shows that inclusion of even this modest mass planet ($0.8 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at ??? AU. Modes of cos($m\phi$) symmetry with $m =$ 5 and 6 dominate. An $m = 3$ (three-armed) spiral also grows somewhat more slowly, but lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominates the GI burst, which also occurs $\sim$ 2 ORPs earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure 2 shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. Beginning at 3 ORP, when the $m = 3$ mode has grown significantly and the planet is lagging one of its arms in azimuth, the planet gains angular momentum and moves outward. At about $t =$ 4 ORP, as the planet overtakes the arm, the sign of the net torque reverses and the planet plunges from 23 to 17 AU in about 4 ORP (10$^3$ yr). During this time, the planet and mode seem to interact strongly enough that the planet essentially surfs along the arm, maintaining a location within the arm as it plunges. After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values are comparably strong. The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at ??? AU in the GI-active disk.

@@

{\it Eccentiricity.} Figure 3 shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases, as it should (ref ???), during the 1.5 ORP when it is migrating only due to its own wake. However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on a short time scale.

{\it Migration Relative to the Disk.} The disk surface density distribution does not vary much in the asymptotic phase. So radial motion of the planet in the $t = 10$ planet run is also radial migration relative to the gas. During the burst (see, e.g., Boley et al. 2006), however, the surface density of the disk changes dramatically. To verify that the planet is really migrating relative to the background gas disk in the $t = 0$ run, we have compared the planet's radius $\varpi_p(t)$ with the mass inside that radius, expressed in terms of $m_{\rm cyl}$, the fraction of the disk mass interior to the planet. Between $t = 3.52$ and $3.89$ ORP, when $\varpi_p$ increases modestly, $m_{\rm cyl}$ increases from about 0.50 to 0.53; between 3.89 and 8.47 ORP, when $\varpi_p$ decreases dramatically, $m_{\rm cyl}$ decreases from 0.53 to only 0.30. So, the planet is migrating significantly with respect to the disk mass even while the inner part of the gas disk moves inward during the burst. For a comparison measure of the magnitude of the gas motion, $m_{\rm cyl}$ at 23 AU increases from 0.50 to 0.60 from $t =$ 3.89 to 8.47 ORP.

@@

Check against estimated type I migration without GIs. {\bf Aaron: Can you do this?} Figure 2 suggests that the overall radial drift is not too different, but it is definitely not monotonic and can stall. Further work is needed to determine how long the stalls last.

Generic relationship to Laughlin's work. Just a few sentences -- structure (gaps, embedded masses) affects GI stability and the modes that grow.

Discrete mode interactions with the planet seem important -- Wind up and pitch. Surfing.

@@

Probably not a random walk for this mass and this disk. Need to know dependence on planet mass and initial location.

Magnitude of e.

\section{Summary}

The inclusion of a 1MJup mass in the equilibrium disk initiates GIs and dramatically effects the character of the GIs.
GIs can cause both inward and outward planet migration and can cause rapid migration in the burst phase.
GIs can have a strong effect on the orbital eccentricity of the included planet.

@@

\begin{center}
{\large\bf Figure Captions}
\end{center}

{\bf Figure 1.}  Am plot.

\begin{center}
{\large\bf Figure Captions}
\end{center}

{\bf Figure 2.}  Plot of r(t) for the point mass in two simulations. The red curve shows the planet’s position when included at t = 0, the black curve for the planet inserted at t = 10 ORPs.  The diamonds track the semi-major axis for each 0 to 2π interval. Figure 2 shows the planet radius over time for both the t = 0 simulation (red) and the t = 10 ORPs simulation (black). The diamonds represent the semi-major axis for each 0 to 2π interval. As can be seen in Figure 2, for the t = 0 simulation the planet begins to drift inward, presumably due to Type 1 migration, and then moves inward rapidly once the burst of GI activity begins around 3.5 ORPs.

\begin{center}
{\large\bf Figure Captions}
\end{center}

{\bf Figure 3.}  Plot of the eccentricities for each of the 0 to 2π intervals for the t = 0 (red) simulation and t = 10 ORPs (black).

>>>>

Section 2 below presents our numerical methods and initial conditions.  We describe the simulation results in Section 3, and discussion them in Section 4.

@@

%Scott: A number of things in this paragraph need to verified by further analysis.
{\it The $t = 0$ Planet Run.} {\bf See note in Latex file.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.8 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at ??? AU. Modes of cos($m\phi$) symmetry with $m =$ 5 and 6 dominate. An $m = 3$ (three-armed) spiral also grows somewhat more slowly, but lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominates the GI burst, which also occurs $\sim$ 2 ORPs earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. Beginning at 3 ORP, when the $m = 3$ mode has grown significantly and the planet is lagging one of its arms in azimuth, the planet gains angular momentum and moves outward. At about $t =$ 4 ORP, as the planet overtakes the arm, the sign of the net torque reverses and the planet plunges from 23 to 17 AU in about 4 ORP (10$^3$ yr). During this time, the planet and mode seem to interact strongly enough that the planet essentially surfs along the arm, maintaining a location within the arm as it plunges. After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values are comparably strong. The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at ??? AU in the GI-active disk.

@@

{\it Eccentiricity.} Figure \ref{fig:e} shows the evolution of $e$ in the planet simulations. In both cases, the planets were inserted with an approximate circular velocity computed by adding the interior disk mass to the stellar point mass. The presence of GI activity in the $t = 10$ planet run immediately increases $e$ to $\sim 0.08$. In the $t = 0$ planet run, the modest initial $e$ decreases, as it should (ref ???), during the 1.5 ORP when it is migrating only due to its own wake. However, once there are strong nonlinear interactions with GI modes, $e$ jumps upward. In both runs, it appears that interaction with well-established GI activity leads to eccentricities between 0.02 and 0.1 that vary in a chaotic way between these extremes on orbit period time scales.

{\it Migration Relative to the Disk.} The disk surface density distribution does not vary much in the asymptotic phase, so radial motion of the planet in the $t = 10$ planet run also represents radial migration relative to the gas. During the burst (e.g., Boley et al. 2006), the surface density of the disk changes dramatically. To verify that the planet is really migrating relative to the background gas disk in the $t = 0$ run, we have compared the evolution of the planet's radial position $\varpi_p(t)$ with the mass inside that radius, expressed in terms of $m_{\rm cyl}$, the fraction of the disk mass interior to the planet. Between $t = 3.52$ and $3.89$ ORP, when $\varpi_p$ increases modestly, $m_{\rm cyl}$ increases from about 0.50 to 0.53; between 3.89 and 8.47 ORP, when $\varpi_p$ decreases dramatically, $m_{\rm cyl}$ also decreases dramatically from 0.53 to 0.30. So, the planet is migrating significantly with respect to the disk mass even while the inner part of the gas disk is moving inward during the burst. For a comparison measure of the magnitude of the gas motion, $m_{\rm cyl}$ at 23 AU increases from 0.50 to 0.60 from $t =$ 3.89 to 8.47 ORP, corresponding to an gas disk inflow rate of $\sim 10^{-5} M_{\odot}/{\rm yr}$.

@@

{\bf See notes in Latex file.}
%Scott and Aaron: This section needs a lot of help.

Check against estimated type I migration without GIs. 
%Aaron: Can you do this?} 
Figure \ref{fig:a} suggests that the overall radial drift is not too different, but it is definitely not monotonic and can stall. Further work is needed to determine how long the stalls last.

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities (Li and Lovelace articles abound 2000, 2001). Meschiari \& Laughlin (2008) recently demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs.

Discrete mode interactions with the planet seem important -- Wind up and pitch? Surfing?

@@

Not a simple random walk for this mass and this disk. Need to know dependence on planet mass and initial location.

Magnitude of $e$ is roughly consistent with $c_s/v_K$ and low Mach number spirals.

The inclusion of a 1$M_J$ mass in the equilibrium disk initiates GIs and dramatically effects the character of the GIs. GIs can cause both inward and outward planet migration and can cause rapid migration in the burst phase.
GIs can have a strong effect on the orbital eccentricity of the included planet.

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of Boley et al. (2006). Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The set of red and black curves correspond to the $t = 0$ and $t = 10$ planet runs, respectively. {\bf The radial migration plot is being used as a placeholder.}}
\label{fig:Am}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planeta.pdf}
\caption{Plot of the planet's radial position as a function of time $\varpi_p(t)$ for the two planet simulations. The red and black indicate the curves ($\varpi(t)$), points ($a$), and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively. The diamonds are the approximate semi-major axes $a$ computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text.}
\label{fig:a}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{../Figures/planete.pdf}
\caption{Plot of the eccentricities $e$, computed for each 0 to 2$\phi$ change in azimuth of the planet as explained in the text, as a function of time. The red and black indicate the points ($e$) and horizontal scales ($t$) that correspond to the $t = 0$ and $t = 10$ planet runs, respectively.}
\label{fig:e}
\end{figure}

%\begin{figure}[t]
%\includegraphics{../Figures/name_of_figure.eps}
%\caption{here's the caption}
%\label{fig:name}
%\end{figure}
